Modelling BCG vaccination in the UK

### 2019-07-04 rmds/98-discussion.Rmd
reorganised thesis to work with online gitbook
Former-commit-id: 3fd8c59438bc22ef6b09d75159886a79ea855f4e
pushing fixes
checked accroynm usage

# Discussion {#discussion}

@@

In Chapter \@ref(data) I found that there was some evidence that negative outcomes were more frequent in TB cases not BCG vaccinated than in those that were. I also found that missingness in routine surveillance sources of Tuberculosis data was associated with multiple risk factors. In Chapter \@ref(beneficial-bcg-out) I found supporting evidence that BCG vaccination was associated with reduced all-cause morality with some evidence that this may have been due to reduced TB mortality. I found little evidence for any other association with TB outcomes, after adjusting for confounding. In Chapter \@ref(sutherland) I explored some of the evidence that was used by policy makers to assess the impact of the change in policy. I found that the previous approach had drastically underestimated the amount of uncertainty surrounding the effect estimates. Using newly available data, I also found that ending the scheme was projected to result in greater number of notifications in the UK born than previously thought. These findings were confirmed in Chapter \@ref(direct-eff) where I evaluated the evidence in the surveillance data that the change in policy had impacted Tuberculosis incidence rates in the target populations. However, in this chapter I found that any increase in TB notifications in the UK born was likely far outweighed by reductions in the number of notifications in the non-UK born. Although I was unable to rule out an unrelated policy change. Finally in Chapter \@ref(model-results) I found that the BCG schools' scheme was projected to *placeholder* TB incidence rates compared to both neonatal and no vaccination over a range of time horizons.

@@

This thesis has used multiple methods, and data sources, to explore the impact of the 2005 change in BCG policy. This triangulation allows for more certainty in the findings than if only a single approach had been used. A limitation of the work in this thesis is that all results were based on surveillance data. Surveillance data is subject to multiple bias issues (see Chapter \@ref(data) for details). Ideally multiple different data types would have been used to more effectively triangulate the impact of the change in policy.  However, the data used in this thesis represents the best available. No similarly thorough use of an equivalent data source exists for Tuberculosis. A major strength of the work in this thesis is the attention that has been paid to make it both open and reproducible. Hopefully, this will allow these findings to be more easily validated, and built upon, by others. Finally the work in this thesis generated several peer reviewed tools as a by-product of the main research question. These tools have benefited other research projects.

@@

This thesis has highlighted the trade-off between vaccinating those at school-age and neonates in settings where the waning of BCG effectiveness when given later in life is minimal. Whilst policy makers were previously aware of this trade-off, their was little quantitative evidence exploring it explicitly. Globally, BCG policy does not account for areas where the BCG vaccine may be equally effective regardless of when it is given.[@TheWorldHealthOrganization:2018va] The evidence from this thesis should, therefore, be considered when outlining future BCG policy. In addition, new Tuberculosis vaccines are in development that may be less susceptible to waning effectiveness when given later in life over a greater geographic area.[@Medicine2013] The findings from this thesis may be applicable to these vaccines in areas where the BCG vaccine is currently known to be ineffective. This may mean that these newly developed vaccines may be better targeted at those at school-age, rather than neonates, depending on the duration of protection that they provide. The work from Chapter \@ref(model-development) and Chapter \@ref(model-fitting) may be particularly suitable to adaption for this use case. This thesis has also explored the potential benefits of BCG vaccination on TB outcomes. The evidence of a reduction in all-cause TB mortality in TB cases may add additional weight to the argument that wider vaccination maybe more cost effective than previously thought in low incidence countries. These findings may also be used to drive vaccine uptake as they provide additional incentives for vaccination. Finally, this thesis has shown the that the impact of the BCG policy has varied depending on UK birth status. This may further strengthen the case for varying vaccination policies depending on the country of origin of the target of vaccination policy, and their immediate families country of origin.

@@

Open reproducible research has been a primary focus of this thesis. A version controlled archive of this thesis is available from GitHub^[https://github.com/seabbs/thesis], with a formatted version available on my personal site^[https://www.samabbott.co.uk/thesis]. This thesis relies on data from the Enhanced Tuberculosis Surveillance system and the Labour Force Survey. The cleaning and munging of this data has been standardised as an R package, `tbinenglanddataclean`^[https://www.samabbott.co.uk/tbinenglanddataclean/], and is available for download. All chapters that contain analysis are linked to their own GitHub repositories, each of which is fully reproducible (discounting the raw data which cannot be released due confidentiality reasons). Literate coding was used to link analysis code with documentation using the R tool chain. An R package, `prettypublisher`^[https://github.com/seabbs/prettypublisher], was developed to augment these tools. Where possible open source tooling has been used to provide a working analytical environment for each chapter. All chapters that have been peer reviewed, or are undergoing peer review, have been preprinted. The model developed in Chapter \@ref(model-development) has been released as an R package along with the fitting pipeline developed in Chapter \@ref(model-fitting). Tools used to develop the figures in Chapter \@ref(background), using World Health Organization data, were expanded into an R package (see Chapter \@ref(getTBinR)). Tooling developed alongside this thesis follows open source best practices. See Chapter \@ref(introduction) for full details on the open source projects developed as part of this thesis.

>>>>

```{r}
##Wordcount: 1865
```
# Discussion {#discussion-chapt}

@@

In Chapter \@ref(data) I found that there was some evidence that negative outcomes were more frequent in Tuberculosis (TB) cases not BCG vaccinated than in those that were. I also found that missingness in routine surveillance sources of TB data was associated with multiple risk factors. In Chapter \@ref(beneficial-bcg-out) I found supporting evidence that BCG vaccination was associated with reduced all-cause morality with some evidence that this may have been due to reduced TB mortality. I found little evidence for any other association with TB outcomes, after adjusting for confounding. In Chapter \@ref(sutherland) I explored some of the evidence that was used by policy makers to assess the impact of the change in policy. I found that the previous approach had drastically underestimated the amount of uncertainty surrounding the effect estimates. Using newly available data, I also found that ending the scheme was projected to result in greater number of notifications in the UK born than previously thought. These findings were confirmed in Chapter \@ref(direct-eff) where I evaluated the evidence in the surveillance data that the change in policy had impacted TB incidence rates in the target populations. However, in this chapter I found that any increase in TB notifications in the UK born was likely far outweighed by reductions in the number of notifications in the non-UK born. Although I was unable to rule out an unrelated policy change. Finally in Chapter \@ref(model-results) I found that the BCG schools' scheme was projected to *placeholder* TB incidence rates compared to both neonatal and no vaccination over a range of time horizons.

@@

This thesis has used multiple methods, and data sources, to explore the impact of the 2005 change in BCG policy. This triangulation allows for more certainty in the findings than if only a single approach had been used. A limitation of the work in this thesis is that all results were based on surveillance data. Surveillance data is subject to multiple bias issues (see Chapter \@ref(data) for details). Ideally multiple different data types would have been used to more effectively triangulate the impact of the change in policy.  However, the data used in this thesis represents the best available. No similarly thorough use of an equivalent data source exists for TB. A major strength of the work in this thesis is the attention that has been paid to make it both open and reproducible. Hopefully, this will allow these findings to be more easily validated, and built upon, by others. Finally the work in this thesis generated several peer reviewed tools as a by-product of the main research question. These tools have benefited other research projects.

@@

This thesis has highlighted the trade-off between vaccinating those at school-age and neonates in settings where the waning of BCG effectiveness when given later in life is minimal. Whilst policy makers were previously aware of this trade-off, their was little quantitative evidence exploring it explicitly. Globally, BCG policy does not account for areas where the BCG vaccine may be equally effective regardless of when it is given.[@TheWorldHealthOrganization:2018va] The evidence from this thesis should, therefore, be considered when outlining future BCG policy. In addition, new TB vaccines are in development that may be less susceptible to waning effectiveness when given later in life over a greater geographic area.[@Medicine2013] The findings from this thesis may be applicable to these vaccines in areas where the BCG vaccine is currently known to be ineffective. This may mean that these newly developed vaccines may be better targeted at those at school-age, rather than neonates, depending on the duration of protection that they provide. The work from Chapter \@ref(model-development) and Chapter \@ref(model-fitting) may be particularly suitable to adaption for this use case. This thesis has also explored the potential benefits of BCG vaccination on TB outcomes. The evidence of a reduction in all-cause TB mortality in TB cases may add additional weight to the argument that wider vaccination maybe more cost effective than previously thought in low incidence countries. These findings may also be used to drive vaccine uptake as they provide additional incentives for vaccination. Finally, this thesis has shown the that the impact of the BCG policy has varied depending on UK birth status. This may further strengthen the case for varying vaccination policies depending on the country of origin of the target of vaccination policy, and their immediate families country of origin.

@@

Open reproducible research has been a primary focus of this thesis. A version controlled archive of this thesis is available from GitHub^[https://github.com/seabbs/thesis], with a formatted version available on my personal site^[https://www.samabbott.co.uk/thesis]. This thesis relies on data from the Enhanced TB Surveillance system and the Labour Force Survey. The cleaning and munging of this data has been standardised as an R package, `tbinenglanddataclean`^[https://www.samabbott.co.uk/tbinenglanddataclean/], and is available for download. All chapters that contain analysis are linked to their own GitHub repositories, each of which is fully reproducible (discounting the raw data which cannot be released due confidentiality reasons). Literate coding was used to link analysis code with documentation using the R tool chain. An R package, `prettypublisher`^[https://github.com/seabbs/prettypublisher], was developed to augment these tools. Where possible open source tooling has been used to provide a working analytical environment for each chapter. All chapters that have been peer reviewed, or are undergoing peer review, have been preprinted. The model developed in Chapter \@ref(model-development) has been released as an R package along with the fitting pipeline developed in Chapter \@ref(model-fitting). Tools used to develop the figures in Chapter \@ref(background), using World Health Organization data, were expanded into an R package (see Chapter \@ref(getTBinR)). Tooling developed alongside this thesis follows open source best practices. See Chapter \@ref(introduction) for full details on the open source projects developed as part of this thesis.

### 2019-07-04 rmds/index.Rmd
Ellens comments
checked accroynm usage

In this thesis, a partially observed semi-stochastic compartmental model of Tuberculosis is developed that models demographic processes such as ageing, births and deaths, as well as vaccination and TB treatment (Chapter \@ref(model-development)). Compartmental infectious disease models generally operate by separating a given population into a series of groups, most commonly susceptible, infectious and recovered populations.[@Anderson1991; @Keeling2007] Movements between these groups are then modeled using a series of differential equations. Transmission is modeled using mass action,[@Anderson1991; @Keeling2007] where infected cases are assumed to randomly interact with susceptible individuals. Additional detail can be added to this model by stratifying the population further and adding additional parameters to modify the degree of mixing between populations. Transition between compartments is assumed to be exponential. See [@Anderson1991; @Keeling2007] for a theoretical introduction to infectious disease models and [@King] for implementation details using R.

@@

To understand the impact of BCG vaccination on the epidemiology of tuberculosis in England, and to use this understanding to forecast the future effects of current and historic BCG vaccination policy.

@@

  - To describe the current epidemiology of Tuberculosis in England, in the context of global Tuberculosis epidemiology.
  - To assess some of the statistical modelling evidence used to justify the 2005 change in BCG vaccination policy in the UK.
  - To assess whether there is evidence in routinely-collected surveillance data that BCG vaccination impacts outcomes for TB cases in England.
  - To assess the effects of the 2005 change in vaccination policy on those eligible for vaccination.
  - To develop a parsimonious transmission dynamic model of tuberculosis which captures current, and historic, vaccination policy and reflects our current understanding of TB epidemiology in England.
  - To fit this model using all available data sources.
  - To investigate the effectiveness of universal school-age vs. universal neonatal vs. no vaccination using the previously developed transmission dynamic model.

@@

* **Chapter \@ref(background):** Background information is given on Tuberculosis and the BCG vaccine. This information helps motivate future chapters and is useful for non-subject area experts.

* **Chapter \@ref(getTBinR):** `getTBinR`, an R package that facilitates downloading Tuberculosis relevant data from the World Health Organization and provides functionality for visualising the downloaded data, is introduced. The motivation and context for this package as part of the wider thesis is also outlined.

* **Chapter \@ref(data):** This chapter describes the epidemiology of tuberculosis in England, using routine national data-sets. Focusing on: the impact of missing data; the mechanisms underlying that missing date; seasonal trends; the role of age; UK birth status; BCG status; trends in TB incidence over time; and TB outcomes in England using case rates. These data are used in all subsequent chapters in this thesis.

@@

* **`getTBinR`**: The `getTBinR` R package facilitates downloading the most up-to-date version of multiple Tuberculosis relevant data sources from the World Health Organization, along with the accompanying data dictionaries. It also contains functions to allow easy exploration of the data via searching data dictionaries, summarising key metrics on a regional and global level, and visualising the data in a variety of customisable ways. See Chapter \@ref(getTBinR) for further details. Install from CRAN with `install.packages("getTBinR")` or install the development version from GitHub with `devtools::install_github("seabbs/getTBinR")`. Link: https://www.samabbott.co.uk/getTBinR/

* **`tbinenglanddataclean`**: An R package that contains the functions and documentation required to reproduce all data import and munging used in this thesis. This package provides a workflow to facilitate reproducing all analyses in this thesis and to expedite the work of others using data from the Enhanced Surveillance System (Chapter \@ref(data)). Available from GitHub using `devtools::install_github("seabbs/tbinenglanddataclean")`. Link: https://www.samabbott.co.uk/tbinenglanddataclean/

@@

* **Explore Tuberculosis in England and Wales**: Developed to allow public Public Health England Tuberculosis Notification data to be explored interactively. Key interventions are highlighted and link to trends in Tuberculosis notifications. This app is used in its static form in Chapter \@ref(background). Link: http://seabbs.co.uk/shiny/TB_England_Wales/

@@

* **Introduction to Tuberculosis models**: Developed to allow simple Tuberculosis models to be explored in an interactive session. Inspired by practicals from the Introduction to Tuberculosis, run by TB MAC (http://tb-mac.org/) at the 2017 Union conference. Link: http://seabbs.co.uk/shiny/intro_to_tb_models/

@@

- This chapter provides an introduction to Tuberculosis and the BCG vaccine. It then motivates the remainder of this thesis. 
- An outline of the theoretical framework used throughout this thesis is given.
- The aims and objectives of this thesis are detailed.
- An overview of the chapters is provided.
- Finally the dissemination of this work so far is summarised, broken down into peer reviewed output, preprints, software output, and talks given at academic conferences.

>>>>

In this thesis, a partially observed semi-stochastic compartmental model of TB is developed that models demographic processes such as ageing, births and deaths, as well as vaccination and TB treatment (Chapter \@ref(model-development)). Compartmental infectious disease models generally operate by separating a given population into a series of groups, most commonly susceptible, infectious and recovered populations.[@Anderson1991; @Keeling2007] Movements between these groups are then modeled using a series of differential equations. Transmission is modeled using mass action,[@Anderson1991; @Keeling2007] where infected cases are assumed to randomly interact with susceptible individuals. Additional detail can be added to this model by stratifying the population further and adding additional parameters to modify the degree of mixing between populations. Transition between compartments is assumed to be exponential. See [@Anderson1991; @Keeling2007] for a theoretical introduction to infectious disease models and [@King] for implementation details using R.

@@

To understand the impact of BCG vaccination on the epidemiology of TB in England, and to use this understanding to forecast the future effects of current and historic BCG vaccination policy.

@@

  - To describe the current epidemiology of TB in England, in the context of global TB epidemiology.
  - To assess some of the statistical modelling evidence used to justify the 2005 change in BCG vaccination policy in the UK.
  - To assess whether there is evidence in routinely-collected surveillance data that BCG vaccination impacts outcomes for TB cases in England.
  - To assess the effects of the 2005 change in vaccination policy on those eligible for vaccination.
  - To develop a parsimonious transmission dynamic model of TB which captures current, and historic, vaccination policy and reflects our current understanding of TB epidemiology in England.
  - To fit this model using all available data sources.
  - To investigate the effectiveness of universal school-age vs. universal neonatal vs. no vaccination using the previously developed transmission dynamic model.

@@

* **Chapter \@ref(background):** Background information is given on TB and the BCG vaccine. This information helps motivate future chapters and is useful for non-subject area experts.

* **Chapter \@ref(getTBinR):** `getTBinR`, an R package that facilitates downloading TB relevant data from the World Health Organization and provides functionality for visualising the downloaded data, is introduced. The motivation and context for this package as part of the wider thesis is also outlined.

* **Chapter \@ref(data):** This chapter describes the epidemiology of TB in England, using routine national data-sets. Focusing on: the impact of missing data; the mechanisms underlying that missing date; seasonal trends; the role of age; UK birth status; BCG status; trends in TB incidence over time; and TB outcomes in England using case rates. These data are used in all subsequent chapters in this thesis.

@@

* **`getTBinR`**: The `getTBinR` R package facilitates downloading the most up-to-date version of multiple TB relevant data sources from the World Health Organization, along with the accompanying data dictionaries. It also contains functions to allow easy exploration of the data via searching data dictionaries, summarising key metrics on a regional and global level, and visualising the data in a variety of customisable ways. See Chapter \@ref(getTBinR) for further details. Install from CRAN with `install.packages("getTBinR")` or install the development version from GitHub with `devtools::install_github("seabbs/getTBinR")`. Link: https://www.samabbott.co.uk/getTBinR/

* **`tbinenglanddataclean`**: An R package that contains the functions and documentation required to reproduce all data import and munging used in this thesis. This package provides a workflow to facilitate reproducing all analyses in this thesis and to expedite the work of others using data from the Enhanced Surveillance System (ETS) (Chapter \@ref(data)). Available from GitHub using `devtools::install_github("seabbs/tbinenglanddataclean")`. Link: https://www.samabbott.co.uk/tbinenglanddataclean/

@@

* **Explore Tuberculosis in England and Wales**: Developed to allow public Public Health England TB Notification data to be explored interactively. Key interventions are highlighted and link to trends in TB notifications. This app is used in its static form in Chapter \@ref(background). Link: http://seabbs.co.uk/shiny/TB_England_Wales/

@@

* **Introduction to Tuberculosis models**: Developed to allow simple TB models to be explored in an interactive session. Inspired by practicals from the Introduction to TB, run by TB MAC (http://tb-mac.org/) at the 2017 Union conference. Link: http://seabbs.co.uk/shiny/intro_to_tb_models/

@@

- This chapter provides an introduction to TB and the BCG vaccine. It then motivates the remainder of this thesis. 
- An outline of the theoretical framework used throughout this thesis is given.
- The aims and objectives of this thesis are detailed.
- An overview of the chapters is provided.
- Finally the dissemination of this work so far is summarised, broken down into peer reviewed output, preprints, software output, and talks given at academic conferences.
